CENSE - Perceptual test II

### 2019-02-25 report.tex
proff read 1

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\DeclareMathOperator*{\argmax}{arg\,max}

\pagestyle{fancy}
\begin{document}
\author{F\'elix GONTIER \\ \texttt{felix.gontier@ls2n.fr}}
\title{CENSE - Perceptual test II}
\maketitle

@@

This report presents the results of a perceptual experiment conducted as part of the Cense project~\cite{cense}. From recordings of a sensor network with low information content, this study aims at developing a processing chain for the estimation of pleasantness in urban sound environments.

@@

Very few recorded scenes are fully annotated in terms of backgrounds, event onsets-offsets and sound levels. However, the prediction of source-specific acoustic indicators through deep learning models requires large amounts of data for which the ground truth of these indicators is known. An alternative to the subjective annotation of corpora of sound scenes is the use of fully simulated scenes, where source activity is controlled. Thus, it is important to verify that new simulated scenes generate a perceptual space similar to that of real scenarios, even with low source complexity.\\

Finally, this work aims at finding physical indicators useful for the estimation of pleasantness. Previous studies consistently model pleasantness as a function of perceptual source activity descriptors such as the time of presence. These quantities are somewhat closer to the physical properties of the scene than the more abstract concept of pleasantness. This work thus primarily focuses on the relation of perceived source activity with physical indicators which can be predicted using deep learning architectures.

@@

\begin{itemize}
\item Background sources that are present throughout the whole scene, and are associated with a sound level.
\item Events characterized for each occurence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

The 19 scenes are then replicated using simScene~\cite{simscene} with a database of extracts from recordings of isolated sound sources. This database is constructed from part of the LibriSpeech~\cite{libri} corpus for voices and Freesound (NOTE?) contributions for all other sources. For this experiment, the playback sound level is the same as during recordings, and ranges from 63.9~dB to 79.4~dB.\\

@@

To study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes from new scenarios. Because of the lack of large isolated sources datasets and the previously proposed models of pleasantness, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}. Considering the aim of the study, newly generated scenarios should cover most encounterable real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{jrg}. These statistics include, conditionally to the ambiance:
\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
\item The inter-onset of event occurences in seconds, also in the form of normal distributions.
\end{itemize}
Adjustments are made on the variance of the considered properties to better cover possible scenarios, and voice events EBR, mostly read English recordings, are reduced to enhance realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. For each environment diverse new scenarios are generated by sampling the background and event properties distributions. For each scene the time of presence for the three sources is estimated using indicators proposed in~\cite{gontier}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\

The experiment corpus is thus comprised of a total of 100 scenes.

@@

Test participants are asked to answer 9 questions per sound scene in the form of 11 point (0-10) scales. They are presented in french and translated in this report. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness (\textit{Unpleasant - Pleasant}) - P
\item Liveliness (\textit{Inert, amorphous - Lively, eventful}) - L
\item Overall loudness (\textit{Quiet - Noisy}) - OL
\item Interest (\textit{Boring, uninteresting - Stimulating, interesting}) - I
\item Calmness (\textit{Agitated, chaotic - Calm, peaceful}) - C
\end{itemize}
These quantities are typically studied in the perceptual characterization of sound scenes (CITE). While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
Furthermore, the primary aim of this test is to model pleasantness from source-specific parameters. Previous studies (CITE) show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence,  or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness can be different for background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles (\textit{Very low - Very high}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds (\textit{Never - Continuously}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
The corresponding interface is shown in Figure~\ref{fig:interface}.

@@

Prior to the test, participants are given an introduction and shown the interface before the test to ensure that the quantities are understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 simulated, then to 25 of the 75 simulated. The evaluated simulated scenes are chosen so that each scene is evaluated by the same number of participants. A random pre-generated order is used to attenuate ordering effects, although all participants are first presented with the most quiet then most noisy of the recorded scenes to calibrate their answers. For the remainder of the corpus all participants experience different scene ordering. Participants can only listen to each scene once. They can start answering questions before the end of the sample, but have to answer all questions and listen to the full extract before proceeding to the next.\\

@@

Participants evaluate each scene on criteria coded on discrete scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus, and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.

@@

\subsection{Correspondance between recorded and replicated scenes}

The relevance of the study relies on the perceptual equivalence of recorded and synthetic scenes. To verify this hypothesis, the 6 recorded and corresponding replicated scenes are compared in terms of perceptual responses.\\

Table~\ref{tab:ogrep} shows the mean differences between assessments for the recorded and corresponding replicated scenes. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. This particular test is chosen because while it assumes that distributions are continuous, it is non-parametric. Pleasantness is found to differ between the two groups only for one location, P4, which is explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is significantly louder in the replicated scene than it is in the recording for about half
of its duration. Although to a lesser extent the same issue explains the high difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly. Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\

Overall, most perceptual responses for replicated scenes are close to those for recordings. No major difference is consistently found in a particular scale, and the proposed hypothesis cannot be invalidated.

@@

The perceptual space generated by the five general scales can also be studied and compared to previous studies to further validate the test material. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The generated spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. This confirms that the simulation process generates perceptually realist scenes and can be used for further research.\\

@@

Finally, we study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene. The resulting values are consistent with the literature~\cite{aumond, lavandier}, both between general scales as discussed in the previous section and for the relations to source contributions. Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while here no direct relation is visible. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\

@@

Both models have very similar expressions, and are comparable to the relation proposed in~\cite{ricciardi2014}, of expression (translated from 1-11 scales to 0-10):
\begin{equation}
\hat P_{p3} = 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}
\end{equation}
although in the considered corpus the overall loudness and presence of voices have respectively a higher and lower impact on pleasantness. This is expected given the observed differences in correlations values.\\

@@

From results obtained in the previous section, a perceptual model is assumed where pleasantness is a function of the perceived overall loudness and time of presence of traffic, voices and birds. These parameters are not easily predictable from recordings because of their subjectivity. Thus, there is a need to propose a similar model of pleasantness from physically-based indicators that can either 1) be computed directly from the audio signal, or 2) be predicted using a learned parametric model taking as input the audio signal.

@@

First, indicators can be computed directly from the mixed audio, without needing information about its composition. Typically this includes indicators derived from sound level measurements. For this study the following are computed with 125~ms (fast) measurements using the Matlab ITA-toolbox~\cite{ita}:

@@

Finally, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier} are computed. $\hat T_s(\alpha)$ is based on a binary emergence model using the full-band source emergence $\Delta_s(t) = L_s(t) - L_{\bar{s}}(t)$ and a threshold $\alpha$ optimized on the present dataset:

@@

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments. While this indicator has a simple expression, it does not reflect well the properties of auditory masking resulting in a low optimal threshold $\alpha_{opt} = -19dB$.\\

@@

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness. Although any could be used, the $L_{50}$ appears to be best suited for pleasantness prediction in this study.\\

Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source. Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weaker for other sources. Both proposed estimates of the time of presence $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ show strong relations to their perceptual counterparts with high correlations (r>0.8), though the latter is consistently better. An interesting result is the high correlation between the $L_{10}$ which is interpreted as a measurement of emergent events to the perceived sound level of passing vehicles $L_{T, p}$. If in perceptual models there is a need to include the sound level of traffic events, the $L_{10}$ may be sufficient instead of a source separation and level estimation algorithm.\\

@@

Table~\ref{tab:physm} compares the performance of the two models. Both obtain good estimates of pleasantness for a 10-point scale. Again, the $\hat P_{\varphi 1}$ model optimized on this experiment's data is expectedly better, though less likely to generalize well to new scenes due to its contruction process. It relies on the characteristics of the simulated scenes in terms of isolated samples variability and content, which leads to corpus-specific source contributions to pleasantness. This is particularly important in the case of voices, found in multiple previous studies to have a positive effect. Conversely, the $\hat P_{\varphi 2}$ model uses a perceptual model constructed on a much larger corpus of recorded scenes, and is less prone to these specificity properties.

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\pagestyle{fancy}
\begin{document}
\author{F\'elix GONTIER \\ \texttt{felix.gontier@ls2n.fr}}
\title{CENSE - Perceptual test II}
\maketitle

@@

This report presents the results of a perceptual experiment conducted as part of the Cense project~\cite{cense}. From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of the pleasantness in urban sound environments.

@@

Very few recorded scenes are fully annotated in terms of backgrounds, event onsets-offsets and sound levels. At the same time, the prediction of source-specific acoustic indicators through deep learning models requires large amounts of data for which the ground truth of these indicators is known. An alternative to the subjective annotation of corpora of sound scenes is the use of fully simulated scenes, where source activity is controlled. It is therefore important to verify that new simulated scenes generate a perceptual space similar to that of real scenarios, even with low source complexity.\\

Next, this work aims at finding physical indicators that can be considered for the estimation of pleasantness. Previous studies consistently model pleasantness as a function of perceptual source activity descriptors such as the time of presence. These quantities are somewhat closer to the physical properties of the scene than the more abstract concept of pleasantness. This work thus primarily focuses on the relation of perceived source activity with physical indicators which can be predicted using deep learning architectures.

@@

\begin{itemize}
\item Background sources that are present throughout the whole scene, and are associated with a sound level.
\item Events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

The 19 scenes are then replicated using simScene~\cite{simscene} with a database of extracts from recordings of isolated sound sources. This database is constructed from part of the LibriSpeech~\cite{libri} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources. For this experiment, the playback sound level is the same as during recordings \ml{of the references scenes ?}, and ranges from 63.9~dB to 79.4~dB.\\

@@

To study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes from new scenarios. Because of the lack of large isolated sources datasets, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}. Considering the aim of the study, newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{jrg}. These statistics include, conditionally to the ambiance:
\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
\item The inter-onset of event occurrences in seconds, also in the form of normal distributions.
\end{itemize}
Adjustments are made on the variance of the considered properties to better cover possible scenarios, and voice events EBR, mostly read English recordings, are reduced to improve realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. For each environment diverse new scenarios are generated by sampling the background and event properties distributions. For each scene the time of presence for the three sources is estimated using indicators proposed in~\cite{gontier}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\

The experiment corpus totals 100 scenes.

@@

Test participants are asked to answer 9 questions per sound scene in the form of 11 point (0-10) scales. They are presented in french and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness (\textit{Unpleasant - Pleasant}) - P
\item Liveliness (\textit{Inert, amorphous - Lively, eventful}) - L
\item Overall loudness (\textit{Quiet - Noisy}) - OL
\item Interest (\textit{Boring, uninteresting - Stimulating, interesting}) - I
\item Calmness (\textit{Agitated, chaotic - Calm, peaceful}) - C
\end{itemize}
These quantities are typically studied in the perceptual characterization of sound scenes (CITE). While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
Furthermore, the primary aim of this test is to model pleasantness from source-specific parameters. Previous studies (CITE) show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence,  or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness can be different for background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles (\textit{Very low - Very high}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds (\textit{Never - Continuously}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
The corresponding interface is shown in Figure~\ref{fig:interface}.

@@

Prior to the test, participants are given a short verbal introduction and the interface is introduced before the test to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 simulated, then to 25 of the 75 simulated. The evaluated simulated scenes are chosen so that each scene is evaluated by the same number of participants. A random pre-generated order is used to attenuate ordering effects, although all participants are first presented with the most quiet then most noisy of the recorded scenes to calibrate their answers. For the remainder of the corpus all participants experience different scene ordering. Participants can only listen to each scene once. They can start answering questions before the end of the sample, but have to answer all questions and listen to the full extract before proceeding to the next.\\

@@

Participants evaluate each scene on criteria coded on a discrete scales \ml{how many steps ?}. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus, and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.

@@

\subsection{Correspondence between recorded and replicated scenes}

The relevance of the remaining study relies on the perceptual equivalence of recorded and synthetic scenes. To verify this hypothesis, the 6 recorded and corresponding replicated scenes are compared in terms of perceptual responses.\\

Table~\ref{tab:ogrep} shows the mean differences between assessments for the recorded and corresponding replicated scenes. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. This non parametric test is chosen because it assumes that distributions are continuous. Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half
of its duration. To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, \ml{trop long : but as voices have a less distinct influence on pleasantness this is not reflected directly}. Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\

Overall, most perceptual responses for replicated scenes are close to those for recordings. No major difference is found in a particular scale, and the proposed hypothesis, \ml{that is that ...} cannot be invalidated.

@@

The perceptual space generated by the five general scales can also be studied and compared to previous studies to further validate the test material. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The generated spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. This confirms that the simulation process generates perceptually convincing scenes and can be used for further research.\\

@@

We then study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene. The resulting values are consistent with the literature~\cite{aumond, lavandier}, both between general scales as discussed in the previous section and for the relations to source contributions. Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\

@@

Both models have very similar formal expressions, and are comparable to the relation proposed in~\cite{ricciardi2014}, of expression (translated from 1-11 scales to 0-10):
\begin{equation}
\hat P_{p3} = 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}
\end{equation}.
Though, in the considered corpus the overall loudness and presence of voices have respectively a higher and lower impact on pleasantness. This is expected given the observed differences in correlations values.\\

@@

From results obtained in the previous section, a perceptual model is chosen where pleasantness is a function of the perceived overall loudness and time of presence of traffic, voices and birds. These parameters are not easily predictable from recordings because of their subjectivity. Thus, there is a need to propose a model of pleasantness whose parameters can be estimated from physically-based indicators that can either 1) be computed directly from the audio signal, or 2) be predicted using a learned parametric model taking as input the audio signal.

@@

First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} when the source are active and at which level. Typically this includes indicators derived from sound level measurements. For this study the following are computed with 125~ms (fast) measurements using the Matlab ITA-toolbox~\cite{ita}:

@@

Next, the $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ time of presence approximations proposed in~\cite{gontier} are computed. $\hat T_s(\alpha)$ is based on a binary emergence model using the full-band source emergence $\Delta_s(t) = L_s(t) - L_{\bar{s}}(t)$ and a threshold $\alpha$ optimized on the present dataset:

@@

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments. While this indicator has a simple expression, it is not found to reflect well the properties of auditory masking resulting in a low optimal threshold $\alpha_{opt} = -19dB$.\\

@@

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all measurements of global sound levels correlate well (r>0.9) with the perceived overall loudness. Although any could be used, the $L_{50}$ appears to be best suited for pleasantness prediction in this study \ml{because ??}.\\

Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source. Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weaker for other sources. Both proposed estimates of the time of presence $\hat T_s(\alpha)$ and $\hat T_s(\alpha, \beta)$ show strong relations to their perceptual counterparts with high correlations (r>0.8), though the latter is consistently better. An interesting result is the large correlation between the $L_{10}$ which is interpreted as a measurement of emergent events to the perceived sound level of passing vehicles $L_{T, p}$. In perceptual models there may be a need to include the sound level of traffic events. In this case, the $L_{10}$ may be sufficient instead of a source separation and level estimation algorithm.\\

@@

Table~\ref{tab:physm} compares the performance of the two models. Both obtain good estimates of pleasantness for a 10-point scale. Again, the $\hat P_{\varphi 1}$ model optimized on this experiment's data is expectedly better, though less likely to generalize well to new scenes due to its construction process. It relies on the characteristics of the simulated scenes in terms of isolated samples variability and content, which leads to corpus-specific source contributions to pleasantness. This is particularly important in the case of voices, found in multiple previous studies to have a positive effect. Conversely, the $\hat P_{\varphi 2}$ model uses a perceptual model constructed on a much larger corpus of recorded scenes, and is less prone to these specificity properties.
